Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-07-10 Notes/galaxy_delensing.tex
refs

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be splitted into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Considering currently available survey, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. The CMB experiments of the current generation will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of the instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular LSST can a achieve a delensing performance comparable to a 3G CMB experiment but and completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.

%

\end{abstract}

@@

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a natural value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski15}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,vanengelen14b,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct the expected realization of lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background as LSS tracer \cite{larsen:2016} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{ carron17}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed.

@@

\subsection{CMB lensing potential}
\label{sec:kappaCMB}
We start from the CMB lensing potential.
The lensing kernel $W^{\kappa}$ is:
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$, $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively. \relook{As can be seen in \reffig{kernel} the lensing kernel $W^{\kappa}(z)$ is quite a broad function of redshift peaking around $z=1.8$ but extending to high redshift.}

@@

Finally, we consider the Square Kilometre Array (SKA).
The SKA is a planned radio array that will survey large scale structure primarily by detecting the redshifted neutral hydrogen (HI) 21cm emission line from a large number of galaxies out to high redshift. 
We will assume an intensity mapping survey mode where SKA will measure the large-scale fluctuations of the \textit{integrated} 21cm intensity from many unresolved galaxies.  
We model both the redshift distribution and  bias of radio sources following \cite{namikawa:2016a}.

\relook{In \reffig{kernel}, we compare the kernels $W^{g}(z)$ of all the tracers introduced here to the CMB lensing kernel $W^{\kappa}(z)$.}

@@

Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit to several experimental results.
For the cross-spectra with the CMB lensing or other galaxy tracers, $C_{\ell}^{{\rm CIB}\mbox{-}j}$, we use the single-SED model of~\cite{hall10}.
It corresponds to the kernel:
\begin{align}\label{eq:12}
W^{\text{CIB}}(z) = b_c\ \frac{\chi^2(z)}{H(z)(1+z)^2}\ e^{-\frac{(z-z_c)^2}{2\sigma^2_z} } f_{\nu(1+z)},
\end {align}
for
\begin{equation}
f_{\nu} = 
\begin{cases}
\Big( e^{\frac{h\nu}{kT}} - 1 \Big)^{-1} \nu^{\beta+3} & (\nu \leq v^{\prime}) \\ \Big( e^{\frac{h\nu^{\prime}}{kT}} - 1 \Big)^{-1} \nu^{\prime \beta+3} \Big( \frac{\nu}{\nu^{\prime}} \Big)^{-\alpha} & (\nu > v^{\prime})
\end{cases}
\end{equation}

We place the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$ and we set $T$ = 34K and $\nu^{\prime} \approx$ 4955 GHz.
\relook{\reffig{kernel} shows how the CIB kernel peaks at higher redshift compared to other galaxy survey kernels, with a better overlap with the lensing one.}

@@

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3. 
SKA in its phase 1 will remove 53\% of the lensing power increasing

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha(r=0), N^{B}_{\ell}=0$ & $\alpha(r=0)$ & $\alpha(r=0.12)$ \\ \hline \hline
LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S4}
\end{table}

@@

Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-survey will remove 56\% of the lensing power. While this performance is comparable to a Stage-3 CMB it will allow to probe the robustness against systematics of an eventual detection.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) we are delensing to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2016,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but will also play a role in cross-checking these internal biases. 
% ===================
\onecolumngrid
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=1]{../images/errors_summary.pdf}
\caption{\relook{to be filled with final results}}
\label{fig}
\end{center}
\end{figure}
\twocolumngrid

% ===================

@@

\bibliographystyle{plain}
\bibliography{/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}

>>>>

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$). We find that, in general, galaxy surveys should be splitted into tomographic bins as this can improve the correlation with CMB lensing by 30\%. 
Considering currently available survey, a DES-like galaxy survey can remove about $14\%$ of the lensing signal. The CMB experiments of the current generation will particularly benefit from large scale structure tracers that, once properly combined, will have a better performance than a CMB internal reconstruction.
With the decrease of the instrumental noise, the CMB internal reconstruction will increase its efficiency and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level.  
Nevertheless optical galaxy surveys will still play an important role even for CMB S4. In particular LSST can a achieve a delensing performance comparable to a 3G CMB experiment but and completely different systematics. This will be important to prove the robustness against systematics of an eventual detection of primordial B-modes.
% maybe non linear scales?
\end{abstract}

@@

Because the CMB B-modes provide the cleanest known observational window into the primordial gravitational wave background, improving their measurement is a major objective of current and future CMB experiments.
Even if inflationary B-modes have not been detected yet, a natural value $r \gtrsim 10^{-3}$, should be reachable shortly given the level of noise expected in future CMB experiments \citep{kamionkowski15}.
However, just reducing the level of noise will not be enough to attain this goal.
Indeed, the observed B modes are not solely sourced by early universe physics; they are also produced by secondary effects taking place in the late-time low-redshift universe. 
In this work, we will focus on the effect of gravitational interactions with large scale structures (LSS).
Lensing shears the CMB polarization pattern, producing ``lensing B modes'' from CMB E modes \citep{zaldarriaga98}.
This expected component has now been measured both in cross-correlation with LSS \citep{hanson13,polarbear2014c,van-engelen:2015,planck2015XV} and from CMB data alone \citep{polarbear2014b,bicep2a,keisler15}.

@@

The optimal way to reduce the lensing contributions to the B-modes is to reconstruct the expected realization of lensing B modes on the observed part of the sky and then use it to clean the data in a process called ``delensing''. 
We can delens the observed B-modes by combining CMB polarization data (what is \textit{lensed}) with tracers of the large scale structure (what is \textit{lensing}) to reconstruct a template of the expected lensing B-modes.
Delensing has been studied for many years \cite{knox2002, kesden2002, seljak2003, simard:2015,sherwin15,smith:2012}.
Furthermore, it has recently been performed on CMB temperature data using the cosmic infrared background as LSS tracer \cite{larsen:2016a} and on CMB temperature and polarization data using CMB data to internally reconstruct the LSS lensing potential \cite{carron:2017}. Finally, the highest B-mode delensing efficiency has been achieved with SPT and Herschel data in which $~28\%$ of the lensing power was removed.

@@

\subsection{CMB lensing potential}
\label{sec:kappaCMB}
We start from the CMB lensing potential.
The lensing kernel $W^{\kappa}$ is:
\begin{equation}
W^{\kappa}(z) = \frac{3\Omega_{\rm m}}{2c}\frac{H_0^2}{H(z)}(1+z)\chi(z)\frac{\chi_*-\chi(z)}{\chi_*},
\end{equation}
%
where $\chi_*$ is the comoving distance to the last-scattering surface at $z_*\simeq 1090$, $\Omega_{\rm m}$ and $H_0$ are the present day values of the Hubble and matter density parameters, respectively. As can be seen in \reffig{kernel}, the lensing kernel $W^{\kappa}(z)$ is a broad function of redshift peaking around $z=1.8$ but extending to high redshift.

@@

Finally, we consider the Square Kilometre Array (SKA).
The SKA is a planned radio array that will survey large scale structure primarily by detecting the redshifted neutral hydrogen (HI) 21cm emission line from a large number of galaxies out to high redshift. 
We will assume an intensity mapping survey mode where SKA will measure the large-scale fluctuations of the \textit{integrated} 21cm intensity from many unresolved galaxies.  
We model both the redshift distribution and  bias of radio sources following \cite{namikawa:2016a}. 
In \reffig{kernel}, we compare the CMB lensing kernel $W^{\kappa}(z)$ to the kernels $W^{g}(z)$ of all the tracers introduced here.

@@

Following \cite{Addison:2011se}, we model the CIB power directly as $C_{\ell}^{{\rm CIB}\mbox{-}{\rm CIB}} = 3500 (l/3000)^{-1.25} {\rm Jy^2 / sr}$.
This model provides an accurate fit to several experimental results.
For the cross-spectra with the CMB lensing or other galaxy tracers, $C_{\ell}^{{\rm CIB}\mbox{-}j}$, we use the single-SED model of~\cite{hall:2010}.
It corresponds to the kernel:
\begin{align}\label{eq:12}
W^{\text{CIB}}(z) = b_c\ \frac{\chi^2(z)}{H(z)(1+z)^2}\ e^{-\frac{(z-z_c)^2}{2\sigma^2_z} } f_{\nu(1+z)},
\end {align}
for
\begin{equation}
f_{\nu} = 
\begin{cases}
\Big( e^{\frac{h\nu}{kT}} - 1 \Big)^{-1} \nu^{\beta+3} & (\nu \leq v^{\prime}) \\ \Big( e^{\frac{h\nu^{\prime}}{kT}} - 1 \Big)^{-1} \nu^{\prime \beta+3} \Big( \frac{\nu}{\nu^{\prime}} \Big)^{-\alpha} & (\nu > v^{\prime})
\end{cases}
\end{equation}

We place the peak of the CIB emissivity at redshift $z_c = 2$ with a broad redshift kernel of width $\sigma_z = 2$ and we set $T$ = 34K and $\nu^{\prime} \approx$ 4955 GHz.
\reffig{kernel} shows how the CIB kernel peaks at higher redshift compared to other galaxy survey kernels, with a better overlap with the CMB lensing one.

@@

\reftab{S4} shows the improvement on the inflationary constraint given the cross-correlation factors described above. 
LSST alone will be able to remove almost half of the B-mode lensing power. This will lead to an improvement in $\sigma(r)$ of a factor of 1.8. If we add all the other LSS tracers previously included (i.e. excluding SKA) the improvement factor increase to 2.3. 
SKA in its phase 1 will remove 53\% of the lensing power increasing 
\relook{FINISH}

@@

\begin{table}
\caption{$\alpha$: Gen 4 experiments}
  \begin{tabular}{|c | c | c | c|}
\hline
Surveys & $\alpha_{r=0}, N^{B}_{\ell}=0$ & $\alpha_{r=0}$ & $\alpha_{r=0.12}$ \\ \hline \hline
LSST (10\%) &1.12 & 1.08  & 1.03 \\ \hline
SKA (0.1$\mu$Jy) (40\%) &1.73& 1.42  & 1.2\\ \hline
CMB S4 (56\%) &2& 1.57  & 1.21 \\ \hline
LSS+CMB (69\%)& 2.8 & 1.8& 1.31  \\ \hline
\end{tabular}
\label{tab:S4_2}
\end{table}

@@

Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-survey will remove 56\% of the lensing power. While this performance is comparable to a Stage-3 CMB it will allow to probe the robustness against systematics of an eventual detection.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) we are delensing to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2017,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but will also play a role in cross-checking these internal biases.

@@

% ===================
\newpage
\onecolumngrid
\begin{figure}[htbp]
%\begin{center}
\includegraphics[scale=1]{../images/errors_summary.pdf}
\caption{\relook{to be filled with final results}}
\label{fig}
%\end{center}
\end{figure}
\twocolumngrid
% ===================

\bibliographystyle{plain}
\bibliography{/Users/alessandromanzotti/Work/Astrophysics/latex_bib/cosmobib,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/2016/delens100d/delens100d,/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
%/Users/alessandromanzotti/Work/Astrophysics/delensing/sptpol_papers/BIBTEX/spt}
